---
title: "tutorial 2 - GWAS Analysis"
output: html
---

##### In this practical we will carry out a complete genome-wide association study (GWAS) on a small example dataset containing about 4,000 individuals and over 300K SNPs using PLINK2. PLINK2 is a widely used, open-source toolkit for GWAS and basic genetic data management. You should have **PLINK2** installed on your machine; please follow the OS-specific setup guide in [`SETUP.md`](../../SETUP.md).

### STEP 0

We will work with simulated case–control data (2,000 cases and 2,000 controls) described in the following papers:

<https://www.nature.com/articles/nprot.2010.116>

<https://www.nature.com/articles/nprot.2010.182>

Please download the three PLINK binary files: `gwa.bed`, `gwa.bim`, and `gwa.fam` from the following Google Drive link and place them in `02_activities/data/`: <https://drive.google.com/drive/folders/1EE7syaxoXmh9wKzJVMY0GfA2os3BF0Xp?usp=drive_link>

### STEP 1

#### Quality Control

In this step we perform basic QC: we remove genotyped SNPs and individuals with too much missing data, and we only keep common variants (MAF ≥ 0.05).

```{r setup, include=FALSE}
# Input files are expected in ./02_activities/data/.
#knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath("../../"))
library(data.table)
library(qqman)
```

```{bash,results='hide',message=FALSE, warning=FALSE}
plink2 --bfile ./02_activities/data/gwa --maf 0.05 --geno 0.01 --mind 0.1 --hwe 0.00005 --make-bed --out ./02_activities/data/gwa.qc
```

Explanation of the main arguments:

-   **maf** – keeps SNPs with minor allele frequency strictly greater than 0.05. Given the small sample size, we focus on common variants.

-   **geno** – excludes SNPs with ≥1% missing genotypes.

-   **mind** – removes individuals whose genotype missingness is ≥10%.

-   **hwe** – removes markers that strongly deviate from Hardy–Weinberg equilibrium (p-value \< 5×10⁻⁵).

**OUTPUT FILES**: `gwa.qc.bed`, `gwa.qc.bim` and `gwa.qc.fam`

#### LD Pruning

We next create a subset of approximately independent SNPs. This LD-pruned set will be used for PCA and removal of related individuals.

```{bash,results='hide',message=FALSE, warning=FALSE}
plink2 --bfile ./02_activities/data/gwa.qc --indep-pairwise 500 50 0.05 --out ./02_activities/data/gwa.qc
```

The three numbers supplied to --indep-pairwise are:

-   Window size in SNPs (here, 500).

-   Step size (the window is shifted by 50 SNPs at a time).

-   LD threshold (r²) (0.05 in this example).

Within each window PLINK computes LD for all SNP pairs, removes one SNP from any pair with r² above the chosen threshold, then slides the window and repeats the process.

We then build a new dataset using only the pruned SNPs:

```{bash,results='hide',message=FALSE, warning=FALSE}
plink2 --bfile ./02_activities/data/gwa.qc --extract ./02_activities/data/gwa.qc.prune.in --make-bed --out ./02_activities/data/gwa.qc_pruned
```


**OUTPUT FILES**: `gwa.qc_pruned.bed`, `gwa.qc_pruned.bim` and `gwa.qc_pruned.fam`

#### Detecting related or duplicate samples

Because the linear regression models used later assume that samples are independent, it is good practice to remove close relatives or duplicate individuals before running a GWAS.

PLINK2 implements the KING-robust kinship method internally. We can use the --king-cutoff option on the LD-pruned dataset to identify a set of unrelated individuals. A cutoff of 0.0884 corresponds roughly to removing up to 2nd-degree relatives.

duplicate / MZ twin: ~0.5

1st degree: ~0.25

2nd degree: ~0.125

3rd degree: ~0.0625

```{bash,results='hide',message=FALSE, warning=FALSE}
plink2 --bfile ./02_activities/data/gwa.qc_pruned --king-cutoff 0.0884 --out ./02_activities/data/gwa_qc
```

--king-cutoff T means: remove one individual from any pair with kinship ≥ T

This command uses the pruned SNP set to estimate pairwise kinship and then automatically chooses a subset of individuals that are mutually unrelated up to the specified cutoff. It produces ID files such as:

`./02_activities/data/gwa_qc.king.cutoff.in.id` – IDs of individuals to keep (unrelated set)

`./02_activities/data/gwa_qc.king.cutoff.out.id` – IDs of individuals to remove

We then create a dataset of unrelated samples:

```{bash,results='hide',message=FALSE, warning=FALSE}
plink2 --bfile ./02_activities/data/gwa.qc --keep ./02_activities/data/gwa_qc.king.cutoff.in.id --make-bed --out ./02_activities/data/gwa.qc_unrelated
```

#### Principal-component analysis (PCA)

To account for population structure, we compute principal components from pruned SNPs in the unrelated dataset.

We reuse the same `./02_activities/data/gwa.qc.prune.in` file to keep the same set of approximately independent SNPs:

```{bash,results='hide',message=FALSE, warning=FALSE}
plink2 --bfile ./02_activities/data/gwa.qc_unrelated --extract ./02_activities/data/gwa.qc.prune.in --make-bed --out ./02_activities/data/gwa.qc_unrelated_pruned
```

Now perform PCA:

```{bash,results='hide',message=FALSE, warning=FALSE}
plink2 --bfile ./02_activities/data/gwa.qc_unrelated_pruned --pca 20 --out ./02_activities/data/gwa_unrel_PCA
```

`./02_activities/data/gwa_unrel_PCA.eigenvec` – individual-level principal components (by default, the top 20 PCs).

`./02_activities/data/gwa_unrel_PCA.eigenval` – eigenvalues corresponding to each component.

These PCs represent major axes of genetic variation (ancestry) in the dataset. More information on PCA with PLINK can be found here: <https://speciationgenomics.github.io/pca/>

### Step 2

#### Logistic regression GWAS

Once QC, pruning, PCA, and relatedness filtering are complete, we can run the association test. We will fit a logistic regression model where:

**Phenotype (case/control) \~ SNP dosage + PC1 + PC2 + PC3**

You can later experiment with including more PCs to see how the results change.

```{bash,results='hide',message=FALSE, warning=FALSE}
plink2 --bfile ./02_activities/data/gwa.qc_unrelated --ci 0.95 --logistic --covar ./02_activities/data/gwa_unrel_PCA.eigenvec --covar-name PC1,PC2,PC3 --adjust --out ./02_activities/data/gwa.qc_assoc
```

Explanation of the main options:

-   **--logistic** : performs SNP-wise logistic regression.

-   **--covar ./02_activities/data/gwa_unrel_PCA.eigenvec** : uses the PCA output (PC scores for each individual) as a covariate file.

-   **--covar-name PC1,PC2,PC3** : includes PC1, PC2, and PC3 from that file as covariates.

-   **--ci 0.95** : requests 95% confidence intervals for effect sizes.

-   **--adjust** : produces multiple-testing adjusted p-values.

With PLINK2, the main association results are written to: `./02_activities/data/gwa.qc_assoc.PHENO1.glm.logistic`

#### Manhattan and QQ plot in R

To visualise the GWAS output, we use the qqman package to create Manhattan and Q–Q plots.

```{r}

#install.packages("data.table")
#install.packages("qqman")

# Load PLINK2 logistic results
assoc <- data.table::fread("./02_activities/data/gwa.qc_assoc.PHENO1.glm.logistic.hybrid")

head(assoc)

assoc_adjusted <- data.table::fread("./02_activities/data/gwa.qc_assoc.PHENO1.glm.logistic.hybrid.adjusted")
head(assoc_adjusted)
```

```{r}

# Keep only the additive test (one row per SNP)
assoc_add <- assoc[TEST == "ADD"]

# qqman expects columns named CHR, BP, SNP, and P.
# In PLINK2 output they are #CHROM, POS, ID, and P.
setnames(assoc_add, c("#CHROM","POS","ID"), c("CHR","BP","SNP"))

# Manhattan plot
png("./02_activities/data/manhattan_plot.png", width = 1200, height = 800, res = 150)

manhattan(assoc_add,
          chr = "CHR",
          bp  = "BP",
         snp = "SNP",
          p   = "P",
          xlab = "",
          ylab = "",
          suggestiveline = FALSE,
          cex.axis = 1.5,
          col = c("lightblue", "lightslateblue"),
          annotatePval=5e-5) 
dev.off()
```

```{r}
# Q-Q plot
png("./02_activities/data/QQ_plot.png", width = 1200, height = 800, res = 150)
qq(assoc_add$P, main = "Q-Q plot of GWAS p-values")
dev.off()
```
#### Generate interactive versions of these plots using the plotly package!

First, install and load the plotly and beepr packages. The beepr package is optional and plays a sound when the plot is ready, which can be helpful for knowing whether a large chunk of code has finished running

```{r}
install.packages("plotly") # for interactive plotting
install.packages("beepr") # for a (fun) notification sound when the plot is ready

library(plotly)
library(beepr) 
```

Interactive Manhattan Plot

```{r}
# assoc_add already exists on this tutorial file from the previous code chunk
# assoc_add <- assoc[TEST == "ADD"]
# setnames(assoc_add, c("#CHROM","POS","ID"), c("CHR","BP","SNP"))

dt <- copy(assoc_add) # create a copy of the data.table to avoid modifying the original

# Ensure numeric chromosome and position (drop non-autosomes)
dt <- dt[!is.na(as.numeric(CHR)) & CHR %in% 1:22]
dt[, CHR := as.numeric(CHR)]
dt[, BP := as.numeric(BP)]

# -log10 p-value for plotting
dt[, logP := -log10(P)]

# Compute cumulative position for x-axis so chromosomes are end-to-end on the x-axis
setorder(dt, CHR, BP) # sort by chromosome and position

chr_sizes <- dt[, .(chr_len = max(BP, na.rm = TRUE)), by = CHR][order(CHR)]
chr_sizes[, chr_start := shift(cumsum(chr_len), fill = 0)]
dt <- merge(dt, chr_sizes[, .(CHR, chr_start)], by = "CHR", all.x = TRUE)

dt[, BP_cum := BP + chr_start]

# For nice x-axis labels at chromosome centers
chr_centers <- chr_sizes[, .(center = chr_start + chr_len / 2), by = CHR]

# Tooltip text with SNP info for top hits (ordered by p-value)
dt[, tooltip := paste0(
  "SNP: ", SNP,
  "<br>CHR: ", CHR,
  "<br>BP: ", BP,
  "<br><i>p</i>: ", signif(P, 3)
)]

# Interactive Manhattan plot
plot_ly(
  data = dt,
  x = ~BP_cum,
  y = ~logP,
  type = "scatter",
  mode = "markers",
  text = ~tooltip,
  hoverinfo = "text",
  marker = list(size = 4),
  color = ~factor(CHR),
  colors = c("lightpink", "lightgreen", "lightslateblue")
) %>%
  layout(
    xaxis = list(
      title = "",
      tickmode = "array",
      tickvals = chr_centers$center,
      ticktext = chr_centers$CHR
    ),
    yaxis = list(title = "-log<sub>10</sub>(<i>p</i>)"),
    legend = list(title = list(text = "CHR")),
 # suggestive line at genome-wide significance threshold (p = 5e-8)
      shapes = list(
      list(
        type = "line",
        xref = "paper", x0 = 0, x1 = 1,
        yref = "y",     y0 = -log10(5e-8), y1 = -log10(5e-8),
        line = list(color = "red", width = 1, dash = "dash")
      )
    )
  )

beep(3) # plays 'fanfare' sound from Final Fantasy when plot is ready
```
Interactive Q-Q Plot
```{r}
# assoc_add already exists on this tutorial file from the previous code chunk
qq_data <- data.table(
  observed = -log10(sort(assoc_add$P)),
  expected = -log10(ppoints(nrow(assoc_add)))
)

# Add tooltip text with SNP info for the top hits (ordered by p-value)
qq_data[, tooltip := paste0(
  "SNP: ", assoc_add$SNP[order(assoc_add$P)],
  "<br>CHR: ", assoc_add$CHR[order(assoc_add$P)],
  "<br>Observed <i>p</i>: ", signif(10^(-observed), 3),
  "<br>Expected <i>p</i>: ", signif(10^(-expected), 3)
)]

# Interactive Q-Q plot
plot_ly(
  data = qq_data,
  x = ~expected,
  y = ~observed,
  type = "scatter",
  mode = "markers",
  text = ~tooltip,
  hoverinfo = "text",
  marker = list(size = 4)
) %>%
  layout(
    xaxis = list(title = "Expected -log<sub>10</sub>(<i>p</i>)"),
    yaxis = list(title = "Observed -log<sub>10</sub>(<i>p</i>)"),
    shapes = list(
      list(
        type = "line",
        xref = "x", x0 = 0, x1 = max(qq_data$expected),
        yref = "y", y0 = 0, y1 = max(qq_data$expected),
        line = list(color = "red", width = 1, dash = "solid")
      )
    )
  )

beep(8) # plays 'mario' sound when plot is ready
```

For more examples and plotting options, see the qqman vignette: <https://cran.r-project.org/web/packages/qqman/vignettes/qqman.html>


#### *Conditional Analysis
```{bash}

plink2 \
  --bfile ./02_activities/data/gwa.qc_unrelated \
  --logistic \
  --covar ./02_activities/data/gwa_unrel_PCA.eigenvec \
  --covar-name PC1,PC2,PC3 \
  --condition rs6768587 \
  --out ./02_activities/data/gwa.qc_assoc_cond_rs6768587
```

```{r}
assoc_condition <- data.table::fread("./02_activities/data/gwa.qc_assoc_cond_rs6768587.PHENO1.glm.logistic.hybrid")

assoc_condition=na.omit(assoc_condition)
head(assoc_condition)

assoc_add1 <- assoc_condition[TEST == "ADD"]
head(assoc_add1)
```

```{r}

# qqman expects columns named CHR, BP, SNP, and P.
# In PLINK2 output they are #CHROM, POS, ID, and P.
setnames(assoc_add1, c("#CHROM","POS","ID"), c("CHR","BP","SNP"))

# Manhattan plot
png("./02_activities/data/manhattan_condition_plot.png", width = 1200, height = 800, res = 150)

 manhattan(assoc_add1,
          chr = "CHR",
          bp  = "BP",
          snp = "SNP",
          p   = "P",
          xlab = "",
          ylab = "",
          suggestiveline = FALSE,
          cex.axis = 1.5,
          col = c("lightblue", "lightslateblue"), annotatePval=5e-5)         
dev.off()
```



 You can also create a text file with one SNP ID per line, for example condition_snps.txt:

rs6768587

rs1234567

rs7654321

```{bash}

setopt interactivecomments

#plink2 \
#  --bfile ./02_activities/data/gwa.qc_unrelated \
#  --logistic \
#  --covar ./02_activities/data/gwa_unrel_PCA.eigenvec \
#  --covar-name PC1,PC2,PC3 \
#  --condition-list condition_snps.txt \
#  --out ./02_activities/data/gwa.qc_assoc_cond_list
```

#### Appendix – Overview of commonly used genetic analysis tools

Below is a short description of several tools that are frequently used in GWAS and related analyses:

[PLINK]{.underline} – Classic GWAS toolkit for QC, data management, and many single-variant association tests.

[PLINK2]{.underline} – Next-generation version of PLINK with extended functionality and improved performance.

[Bcftools]{.underline} – Command-line programs for viewing, filtering, and manipulating VCF/BCF variant files.

[VCFtools]{.underline} – Suite of utilities for summarising, filtering, and converting VCF data.

[REGENIE]{.underline} – Fast whole-genome regression framework for single-variant and gene-based tests in large cohorts.

[SAIGE]{.underline} – Mixed-model method that accommodates unbalanced case–control ratios and large sample sizes.

[GCTA]{.underline} – Software for estimating SNP-heritability, genetic correlation, and performing related analyses.

[KING]{.underline} – Toolkit for inferring kinship, detecting sample duplicates, and exploring population structure.

[FastPCA]{.underline} – Efficient algorithms to compute principal components on large genotype matrices.

[STRUCTURE]{.underline} – Bayesian clustering approach for inferring population structure and assigning individuals to populations.

[ADMIXTURE]{.underline} – Fast maximum-likelihood method for estimating ancestry proportions in admixed individuals.

[ANNOVAR]{.underline} – Framework for functionally annotating genetic variants.

[GATK]{.underline} – Broad Institute toolkit for variant calling and processing of high-throughput sequencing data.

[TensorQTL]{.underline} – GPU-optimised software for rapid QTL mapping.
