---
title: "Assignment #2"
format: html
---

You will need **PLINK2** installed and available in your PATH. Please follow the OS-specific setup guide in [`SETUP.md`](../../SETUP.md). The dataset for this assignment consists of the following binary PLINK files: `gwa.A2.bed`, `gwa.A2.bim`, `gwa.A2.fam` , available at the following Google Drive link: <https://drive.google.com/drive/folders/1rHoy3z52Yyj985ukjjtLhfIBxchpyYtZ?usp=drive_link>. Please download all three files and save them in `02_activities/data/`.

#### Question 1: Data inspection

Before you run any models, first get familiar with the dataset. You may find `data.table::fread()` in R helpful for reading `.bim` and `.fam` files.

(i) Read the `.fam` file. How many samples does the dataset contain?

Load the necessary libraries and set the working directory to the root of the repository 
```{r setup, include=FALSE}
# Input files are expected in ./02_activities/data/.
#knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath("../../"))
library(data.table)
library(qqman)
library(ggplot2)
library(plotly)
library(beepr)
```

Conduct sample number inspection
```{r}
famA2 <- data.table::fread("02_activities/data/gwa.A2.fam")
n_samplesA2 <- nrow(famA2)
n_samplesA2 

# n_samplesA2 corresponds to the number of rows in the .fam file, 4000 in this case
```
Inspect the phenotype to find out how it is distributed
```{r}
setnames(famA2, c("FID", "IID", "PID", "MID", "SEX", "PHENO")) # there are on -9s

ggplot(famA2, aes(x = PHENO)) +
  geom_histogram(binwidth = 1, fill = "lightslateblue", color = "black") +
  labs(title = "Assignment 2 phenotype distribution", x = "Phenotype", y = "Count") +
  theme_minimal()

# From the histogram, the phenotype is continuous and is approximately normally distributed
```

(ii) Read the `.bim` file. How many SNPs does the dataset contain?

```{r}
bimA2 <- data.table::fread("02_activities/data/gwa.A2.bim")
n_snpA2 <- nrow(bimA2)
n_snpA2 

# n_snpA2 corresponds to the number of rows in the .bim file, 306102 in this case
```

#### Question 2: Quality control (QC)

Now we will perform QC using PLINK2 for the genotype files in `gwa.A2`.

(i) Using PLINK2 from the command line (bash), perform basic QC with the following filters: MAF ≥ 0.05, SNP missingness (`--geno`) ≤ 0.01, individual missingness (`--mind`) ≤ 0.10, and HWE p-value ≥ 0.00005, and output the QC’ed dataset as `gwa.qc.A2`.

```{bash}
plink2 --bfile 02_activities/data/gwa.A2 --maf 0.05 --geno 0.01 --mind 0.10 --hwe 0.00005 --make-bed --out 02_activities/data/gwa.qc.A2
```

#### Question 3: Relatedness

In this question, you will use **PLINK2’s built-in KING-robust kinship** (`--king-cutoff`) detect and remove related individuals.

i)  Perform LD pruning on `gwa.qc.A2` using PLINK2 with the following parameters: `--indep-pairwise 500 50 0.05`, and then generate a new dataset containing only the pruned SNPs.

```{bash}
plink2 --bfile 02_activities/data/gwa.qc.A2 --indep-pairwise 500 50 0.05 --out 02_activities/data/gwa.qc.A2
```

(ii) Use PLINK2 on the LD-pruned dataset to identify a set of unrelated individuals up to (approximately) 2nd-degree relatives (use a kinship cutoff of 0.0884).

```{bash}
plink2 --bfile 02_activities/data/gwa.qc.A2 --extract 02_activities/data/gwa.qc.A2.prune.in --king-cutoff 0.0884 --out 02_activities/data/gwa.qc.A2_pruned
```

(iii) Using the unrelated individual list from part (ii), create a dataset containing only unrelated individuals and name it `gwa.qc_unrelated.A2`.

```{bash}
plink2 --bfile 02_activities/data/gwa.qc.A2 --keep 02_activities/data/gwa_qc.king.cutoff.in.id --make-bed --out 02_activities/data/gwa.qc_unrelated.A2
```

#### Question 4: principal component analysis (PCA)

(i) Using PLINK2, perform PCA on the unrelated, LD-pruned dataset to obtain principal components.

    (Hint: Refer to the PCA section in the GWAS Tutorial II. You should use a `*.prune.in` file to select a set of independent SNPs before performing PCA.)

Select a set of independent (unrelated and pruned) SNPs
```{bash}
plink2 --bfile 02_activities/data/gwa.qc_unrelated.A2 --extract 02_activities/data/gwa.qc.A2.prune.in --make-bed --out 02_activities/data/gwa.qc.A2_unrelated_pruned
```
Perform PCA on the selected SNPs from the step above
```{bash}
plink2 --bfile 02_activities/data/gwa.qc.A2_unrelated_pruned --pca 20 --out 02_activities/data/gwa.A2_unrel_PCA
```

#### Question 5: GWAS analyses

We now test for association between each SNP and the continuous phenotype, adjusting for population structure using the top 3 PCs.

Assume that:

-   You will run GWAS on the unrelated, QC’ed dataset `gwa.qc_unrelated.A2`.

-   You have a covariate file (e.g., the `*.eigenvec` output from the PCA analysis) containing PC1–PC3 for each individual.

(i) Using PLINK2, run a linear regression GWAS using `gwa.qc_unrelated.A2` as the input dataset. Adjust for PCs 1–3 as covariates.

```{bash}
plink2 --bfile 02_activities/data/gwa.qc_unrelated.A2 --ci 0.95 --linear --covar 02_activities/data/gwa.A2_unrel_PCA.eigenvec --covar-name PC1,PC2,PC3 --adjust --out 02_activities/data/gwa.qc.A2_assoc
```

(ii) Create a Manhattan plot of the GWAS results.

Prepare data set for plot creation 
```{r}
# Load the PLINK2 linear regression results
assocA2 <- data.table::fread("./02_activities/data/gwa.qc.A2_assoc.PHENO1.glm.linear")
head(assocA2)

# Keep only the additive test (one row per SNP)
assocA2_add <- assocA2[TEST == "ADD"]

# qqman expects columns named CHR, BP, SNP, and P

# In PLINK2 output they are #CHROM, POS, ID, and P
setnames(assocA2_add, c("#CHROM","POS","ID"), c("CHR","BP","SNP"))
```

Produce still and interactive Manhattan plots
```{r}
# Still Manhattan plot

manhattan(assocA2_add,
          chr = "CHR",
          bp  = "BP",
          snp = "SNP",
          p   = "P",
          main = "Manhattan plot of Assignment 2 GWAS p-values",
          xlab = "Chromosome",
          ylab = expression(-log[10](italic(p))),
          col = c("#4E79A7", "#F28E2B"),
          cex = 0.6,
          cex.axis = 1.2,
          genomewideline = -log10(5e-8),
          suggestiveline = FALSE
          )

# Interactive Manhattan plot

# Data preparation for interactive plot
dtA2 <- copy(assocA2_add) # copy of the data table to avoid modifying the original

# Ensure numeric chromosome and position (drop sex chromosomes)
dtA2 <- dtA2[!is.na(as.numeric(CHR)) & CHR %in% 1:22]
dtA2[, CHR := as.numeric(CHR)]
dtA2[, BP := as.numeric(BP)]

# -log10 p-value for plotting
dtA2[, logP := -log10(P)]

# Compute cumulative position for x-axis so chromosomes are end-to-end on the x-axis
setorder(dtA2, CHR, BP) # sort by chromosome and position

chr_sizesA2 <- dtA2[, .(chr_len = max(BP, na.rm = TRUE)), by = CHR][order(CHR)]
chr_sizesA2[, chr_start := shift(cumsum(chr_len), fill = 0)]
dtA2 <- merge(dtA2, chr_sizesA2[, .(CHR, chr_start)], by = "CHR", all.x = TRUE)

dtA2[, BP_cum := BP + chr_start]

# For x-axis labels at chromosome centers
chr_centersA2 <- chr_sizesA2[, .(center = chr_start + chr_len / 2), by = CHR]

# Tooltip text with SNP info for top hits (ordered by p-value)
dtA2[, tooltip := paste0(
  "SNP: ", SNP,
  "<br>CHR: ", CHR,
  "<br>BP: ", BP,
  "<br><i>p</i>: ", signif(P, 3)
)]

# Define chromosome levels and make a factored version of CHR
chr_levels <- as.character(1:22)
dtA2$CHR_f <- factor(dtA2$CHR, levels = chr_levels)

# Plot 
plot_ly(
  data = dtA2,
  x = ~BP_cum,
  y = ~logP,
  type = "scatter",
  mode = "markers",
  text = ~tooltip,
  hoverinfo = "text",
  marker = list(size = 4),
  color = ~CHR_f,
  colors = rep(c("#4E79A7", "#F28E2B"), length.out = length(chr_levels))
) %>%
  layout(
    title = list(text = "Interactive Manhattan plot of Assignment 2 GWAS p-values"),
    xaxis = list(
      title = "Chromosome",
      tickmode = "array",
      tickvals = chr_centersA2$center,
      ticktext = chr_centersA2$CHR
    ),
    yaxis = list(title = "-log<sub>10</sub>(<i>p</i>)"),
    legend = list(title = list(text = "CHR")),
      shapes = list(
      list(
        type = "line",
        xref = "paper", x0 = 0, x1 = 1,
        yref = "y",     y0 = -log10(5e-8), y1 = -log10(5e-8),
        line = list(color = "red", width = 1, dash = "dash")
      )
    )
  )

beep(5) # 'treasure' sound from Zelda
```

(iii) Create a QQ plot of the GWAS p-values.

Produce still and interactive Q-Q plots
```{r}
# Still Q-Q plot

qq(assocA2_add$P, 
   main = "Q-Q plot of Assignment 2 GWAS p-values",
   col  = "#4E79A7"
   )

# Interactive Q-Q plot

# assocA2_add already exists on this tutorial file from the previous code chunk
qq_dataA2 <- data.table(
                        observed = -log10(sort(assocA2_add$P)),
                        expected = -log10(ppoints(nrow(assocA2_add)))
                        )

# Add tooltip text with SNP info for the top hits (ordered by p-value)
qq_dataA2[, tooltip := paste0(
  "SNP: ", assocA2_add$SNP[order(assocA2_add$P)],
  "<br>CHR: ", assocA2_add$CHR[order(assocA2_add$P)],
  "<br>Observed <i>p</i>: ", signif(10^(-observed), 3),
  "<br>Expected <i>p</i>: ", signif(10^(-expected), 3)
)]

# Plot
plot_ly(
  data = qq_dataA2,
  x = ~expected,
  y = ~observed,
  type = "scatter",
  mode = "markers",
  text = ~tooltip,
  hoverinfo = "text",
  marker = list(size = 4, color = "#4E79A7" )
) %>%
  layout(
    title = list(text = "Interactive Q-Q plot of Assignment 2 GWAS p-values"),
    xaxis = list(title = "Expected -log<sub>10</sub>(<i>p</i>)"),
    yaxis = list(title = "Observed -log<sub>10</sub>(<i>p</i>)"),
    shapes = list(
      list(
        type = "line",
        xref = "x", x0 = 0, x1 = max(qq_dataA2$expected),
        yref = "y", y0 = 0, y1 = max(qq_dataA2$expected),
        line = list(color = "red", width = 1, dash = "solid")
      )
    )
  )

beep(4) # "work complete!"
```

### Criteria

| Criteria | Complete | Incomplete |
|:--------------|:----------------------------------|:----------------------|
| **Data inspection** | Correct sample and SNP counts | Counts or phenotype description/plot missing or incorrect. |
| **QC & LD pruning** | Correct PLINK2 QC command and thresholds. | QC/pruning commands, thresholds, or output datasets missing or incorrect. |
| **Relatedness & PCA** | Correct use of PLINK2 command to obtain unrelated samples and PCA run on pruned SNPs. | Relatedness step, unrelated dataset, or PCA analysis missing or incorrect. |
| **GWAS & visualisation** | Linear regression GWAS with PCs as covariates; Manhattan and QQ plots produced. | GWAS command, or Manhattan/QQ plots missing or clearly incorrect. |

### Submission Information

🚨 **Please review our [Assignment Submission Guide](https://github.com/UofT-DSI/onboarding/blob/main/onboarding_documents/submissions.md)** 🚨 for detailed instructions on how to format, branch, and submit your work. Following these guidelines is crucial for your submissions to be evaluated correctly.

------------------------------------------------------------------------

#### Submission Parameters

-   Submission Due Date: 11:59 PM – 01/04/2026

-   The branch name for your repo should be: `assignment-2`

-   What to submit for this assignment:
    -   Populate this Quarto document (`assignment_2.qmd`).
    -   Render the document with Quarto: `quarto render assignment_2.qmd`.
    -   Submit both `assignment_2.qmd`, the rendered HTML file `assignment_2.html` and saved figures in your pull request.

-   What the pull request link should look like for this assignment: `https://github.com/<your_github_username>/gen_data/pull/<pr_id>`

    -   Open a private window in your browser. Copy and paste the link to your pull request into the address bar. Make sure you can see your pull request properly. This helps the technical facilitator and learning support team review your submission easily.

------------------------------------------------------------------------

Checklist:

-   Create a branch called `assignment-2`.
-   Ensure that your repository is public.
-   Review [the PR description guidelines](https://github.com/UofT-DSI/onboarding/blob/main/onboarding_documents/submissions.md#guidelines-for-pull-request-descriptions) and adhere to them.
-   Verify that your link is accessible in a private browser window.
-   Confirm that both `assignment_2.qmd` and `assignment_2.html` are included in the pull request.
